---
title: "Run multivariate models"
author: "Helen Fielding"
format: html
editor: source
---



Author: Helen Fielding

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r Packages and functions and load data}
library(tidyverse)
library(lme4)
library(broom.mixed)
library(survival)
library(ggsurvfit)

theme_set(theme_minimal())

# function to do predictions from mixed models and mutate cond
source('functions.R')

dat <- read_csv('coxph_death_publish.csv')%>%
  mutate(across(.cols = c(sex, colour, diet_type2, imd_quintile), as.factor))

```

```{r cox12 tt mean bmi }
cox12 <- coxph(Surv(censor_age_days, status_death) ~ 
      sex + household_smoking_status + titbits + diet_type2 + diet_raw +
      colour2 + imd_cat + ovary_exp +
      mean_bmi + tt(mean_bmi) +
      exa_fetch_cat_90,
      tt = function(x, t, ...) x * log((t / 365.25) / 10),# estimates effect at 10 years
      data = dat)

summary(cox12)
```

Checking informative censoring

We will now test the sensitivity of the informative censoring by constructing two scenarios;

### 1. High risk model

Non-administrative (LTFU) subjects are die the day of censoring.

```{r Sensitivity of informative censoring - only LTFU}
# dave collett book
# high risk censoring - assume event occurs just after censoring

dat_high_ltfu <- dat%>%mutate(status_death = ifelse(censor_status == 'ltfu', 1, status_death))# just change those not administrative censoring
                         
high_risk_mod_ltfu <- coxph(Surv(censor_age_days, status_death) ~ 
      sex + household_smoking_status + titbits + diet_type2 + diet_raw +
      colour2 + imd_cat + ovary_exp +
      mean_bmi + tt(mean_bmi) +
      exa_fetch_cat_90,
      tt = function(x, t, ...) x * log((t / 365.25) / 10),
      data = dat_high_ltfu)

summary(high_risk_mod_ltfu)
```

### 2. Low risk model

Low risk censoring - LTFU censored subjects live as long as the longest living subject in the study (`r dat%>%filter(status_death == 1)%>%pull(censor_age_days)%>%max()` days).

```{r low risk model - ltfu }
max_surv <- dat%>%filter(status_death == 1)%>%pull(censor_age_days)%>%max()

dat_low_ltfu <- dat%>%mutate_cond(status_death == 0 & censor_status == 'ltfu', censor_age_days = max_surv, status_death = 1)

low_risk_mod_ltfu <- coxph(Surv(censor_age_days, status_death) ~ 
      sex + household_smoking_status + titbits + diet_type2 + diet_raw +
      colour2 + imd_cat + ovary_exp +
      mean_bmi + tt(mean_bmi) +
      exa_fetch_cat_90,
      tt = function(x, t, ...) x * log((t / 365.25) / 10),
      data = dat_low_ltfu)

summary(low_risk_mod_ltfu)
```

```{r Compare models}

sig_vars_test <- tidy(cox12, conf.int = T, exponentiate = T)%>%mutate(model = 'normal')%>%
  bind_rows(tidy(high_risk_mod_ltfu, conf.int = T, exponentiate = T)%>%mutate(model = 'high risk ltfu'))%>%
  bind_rows(tidy(low_risk_mod_ltfu, conf.int = T, exponentiate = T)%>%mutate(model = 'low risk ltfu'))%>%
  mutate(sig = p.value < 0.05)

sig_vars_test%>%ggplot(aes(x = estimate, y = term, xmin = conf.low, xmax = conf.high, colour = model, shape = sig,
                           width = 0.3))+
  geom_vline(aes(xintercept = 1), linetype = 'dashed')+
  geom_point(position = position_dodge(width = 0.5), alpha = 0.9)+
  geom_errorbar(position = position_dodge(width = 0.5))+
  coord_flip()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Some differences in estimates, suggestive of informative censoring - proceed with weighted model.

## IPCW models 

1.  Create a Cox PH with censoring as the response.

```{r Inverse probability of (LTFU) censoring weight - Dave Collett book}

dat%>%count(censor_status)
dat_cens2 <- dat%>%mutate(status_censor = ifelse(censor_status == 'ltfu', 1,0))

dat_cens2%>%count(status_death, status_censor)

ipcw_weight <- coxph(Surv(censor_age_days, status_censor) ~ sex + household_smoking_status + titbits + diet_type2 + diet_raw +
      colour2 + imd_cat + ovary_exp +
      mean_bmi + 
      exa_fetch_cat_90,
                  data = dat_cens2)

summary(ipcw_weight)
cox.zph(ipcw_weight)

# Step 2: Predicted probabilities
sf_cens <- survfit(ipcw_weight, newdata = dat_cens2)
# Extract survival at each person’s observed time
# (probability of remaining uncensored)
S_uncens_matrix <- summary(sf_cens, times = dat_cens2$censor_age_days, extend=TRUE)$surv
S_uncens <- diag(S_uncens_matrix)
# Step 3: Inverse probability weights
# Inverse probability weights
dat$ipw <- 1 / S_uncens
hist(dat$ipw)


```

Household smoking status violates PH. try to stratify model on smoking.

```{r Stratify model - take a while to run}
censor1 <- coxph(Surv(censor_age_days, status_censor) ~ sex + strata(household_smoking_status) + titbits + diet_type2 + diet_raw +
      colour2 + imd_cat + ovary_exp +
      mean_bmi + 
      exa_fetch_cat_90,
                  data = dat_cens2)

summary(censor1)
cox.zph(censor1)
dat_cens2 %>% count(household_smoking_status, status_censor)

# Step 2: Predicted probabilities
sf_cens <- survfit(censor1, newdata = dat_cens2)

S_uncens <- vapply(seq_len(nrow(dat_cens2)), function(i) {
  summary(sf_cens[i], times = dat_cens2$censor_age_days[i], extend = TRUE)$surv
}, numeric(1))

dat$ipcw_stratified1 <- 1 / S_uncens

# no cap needed as not skewed

summary(dat$ipcw_stratified)
summary(dat$ipcw_stratified1)
```

## Fit the weighted Cox PH model.

```{r Fit weighted model}
# Step 4: Fit weighted Cox proportional hazards model
fit_ipcw <- coxph(Surv(censor_age_days, status_death) ~ 
      sex + household_smoking_status + titbits + diet_type2 + diet_raw +
      colour2 + imd_cat + ovary_exp +
      mean_bmi + tt(mean_bmi) +
      exa_fetch_cat_90,
      tt = function(x, t, ...) x * log((t / 365.25) / 10),# estimates effect at 10 years
      data = dat, weights = ipcw_stratified)

summary(fit_ipcw)
```

```{r Compare hazard ratios between all models}
sig_vars_test <- tidy(cox12, conf.int = T, exponentiate = T)%>%mutate(model = 'Unweighted')%>%
  bind_rows(tidy(fit_ipcw, conf.int = T, exponentiate = T)%>%mutate(model = 'IPCW'))%>%
  bind_rows(tidy(high_risk_mod_ltfu, conf.int = T, exponentiate = T)%>%mutate(model = 'High risk'))%>%
  bind_rows(tidy(low_risk_mod_ltfu, conf.int = T, exponentiate = T)%>%mutate(model = 'Low risk'))%>%
  bind_rows(tidy(ipcw_weight, conf.int = T, exponentiate = T)%>%mutate(model = 'Censoring'))%>%
  mutate(sig = p.value < 0.05)

sig_vars_test%>%ggplot(aes(x = estimate, y = term, xmin = conf.low, xmax = conf.high, colour = model, #shape = sig,
                           width = 0.3))+
  geom_vline(aes(xintercept = 1), linetype = 'dashed')+
  geom_point(position = position_dodge(width = 0.5), alpha = 0.9)+
  geom_errorbar(position = position_dodge(width = 0.5))+
  scale_x_continuous(breaks = seq(0.6, 1.8, 0.2))
```

```{r Merge standardised hazard ratios}

cont_summary_values <- dat%>%
  select(mean_bmi#, ex_fetch_mean_a
         )%>%
  gather(key, value)%>%
  group_by(key)%>%
  summarise_all(list(~mean(.), ~sd(.),
                     q10 = ~quantile(., probs = c(0.10))[[1]],
                     q90 = ~quantile(., probs = c(0.90))[[1]],
                     ~min(.), ~max(.)))%>%
  mutate(df10_90 = q90 -q10)%>%
  select(key, sd, df10_90, q10, q90)

hr_var_mapping <- read_delim(
"Variable\tterm
Sex: Male\tsexmale
Colour: Chocolate\tcolour2chocolate
Colour: Yellow\tcolour2yellow
Gonadal exposure (<= 448 days)\tovary_explow
Mean BMI in 2nd yr (at 10 yrs)\tmean_bmi
Mean BMI in 2nd yr (over time)\ttt(mean_bmi)
Extreme fetch in 2nd yr (>90th PC)\texa_fetch_cat_90
Titbits given\ttitbits
Dry food only\tdiet_type2dry_only
Raw diet\tdiet_raw
Smoking in household\thousehold_smoking_status
IMD category 4-5\timd_cat",
delim = "\t",
show_col_types = FALSE)

hr_df_all <-  tidy(cox12, conf.int = T, exponentiate = F)%>%mutate(Model = 'Unweighted')%>%
  bind_rows(tidy(fit_ipcw, conf.int = T, exponentiate = F)%>%mutate(Model = 'IPCW'))%>%
  bind_rows(tidy(high_risk_mod_ltfu, conf.int = T, exponentiate = F)%>%mutate(Model = 'High risk'))%>%
  bind_rows(tidy(low_risk_mod_ltfu, conf.int = T, exponentiate = F)%>%mutate(Model = 'Low risk'))%>%
  bind_rows(tidy(ipcw_weight, conf.int = T, exponentiate = F)%>%mutate(Model = 'Censoring'))%>%
  mutate(term_short = case_when(
  #grepl('ex_fetch_mean_a', term) ~ 'ex_fetch_mean_a',
    grepl('mean_bmi', term) ~ 'mean_bmi',
  .default = term))%>%
  left_join(cont_summary_values, by = c('term_short' = 'key'))%>%
  mutate(est_adj = ifelse(is.na(df10_90), estimate, estimate*df10_90),
         conf.low_adj = ifelse(is.na(df10_90), conf.low, conf.low*df10_90),
         conf.high_adj = ifelse(is.na(df10_90), conf.high, conf.high*df10_90))%>%
  mutate(across(c(estimate, conf.low, conf.high, est_adj, conf.high_adj, conf.low_adj), exp))%>%
  inner_join(hr_var_mapping, by = "term")%>%
  mutate(across(where(is.numeric), round, 3))%>%
  mutate(Model = factor(Model, levels = c("Unweighted", "IPCW", "subset", "High risk", "Low risk", "Censoring")))

# Save out file for use in clean results script
write_csv(hr_df_all, 'hazard_ratios_all_models_standardised_exp.csv')

```
